Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-12-19 src/tex/ms.tex

\section{Method}
\label{sec:method}

We define the progenitor's zero age main sequence (ZAMS) properties like mass,
orbital period, and eccentricity as progenitor parameters $\bm{\theta'}$, the
parameters that control different uncertain physical prescriptions such as wind
strength and common envelope efficiency as hyperparameters $\bm{\lambda}$, and
random variables that affect certain stochastic process such as whether CO natal
kick unbinds a binary as $\bm{X}$. To avoid clutter in the following derivation,
we denote all the parameters related to mapping a particular progenitor system
into a GW event collectively as evolutionary parameters $\bm{\Theta}$, which
includes $\bm{\theta'}$, $\bm{\lambda}$, and $\bm{X}$.

@@

@@

We apply the method described in section \ref{sec:method} to the first GW event GW150914 \citep{GW150914}.
We use the \texttt{Overall\_posterior} posterior samples for GW150914,
publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}. 
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space $\bm{\theta}$ in both the
root-finding stage and the MCMC stage. For progenitor parameters $\bm{\theta'}$, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.
For hyperparameters $\bm{\lambda}$, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$. For systems which begin a common envelope 
with a Hertzsprung-gap donor, we always assume the pessimistic outcome that the two 
stars merge since the donor has likely not formed a strong core-envelope boundary 
\citep{Ivanova2004, Belczynski2008}.

We show a portion of the joint posterior of the progenitor parameters and
hyperparameters of GW150914 in figure \ref{fig:GW150914_posterior}. Each panel
is colored based on the comparison between progenitor parameters (blue),
hyperparameters (orange), or a mix of the two (green). We find that the choice
for $\sigma$ does not affect the formation of GW150914-like BBHs. This is not
unexpected since we use the \citet{Fryer2012} delayed model which reduces the
natal kick strength at BH formation based on the amount of fallback onto the
proto-CO. In the case of GW150914-like BBHs, the BH progenitors are massive
enough that the fallback reduces the natal kick to zero. Because of this we do
not include $\sigma$ in Figure~\ref{fig:GW150914_posterior}. Similarly, we do
not include the ZAMS orbital period and eccentricity, which are correlated with
one another, but not strongly with the other progenitor parameters or
hyperparameters. We find strong correlations between the ZAMS masses, as well as
strong correlations between $M_{2,\rm{ZAMS}}$ and $f_{\rm{acc}}$, and
$f_{\rm{acc}}$ and $\alpha$. We show scatter plots of each of these combinations
in the upper-right inset of the figure.

>>>>

\section{Methods}
\label{sec:method}

\subsection{\textbf{Statistical method}}
\label{subsec:stats}
We define the progenitor's zero age main sequence (ZAMS) properties like mass,
orbital period, and eccentricity as progenitor parameters $\bm{\theta'}$, the
parameters that control different uncertain physical prescriptions such as wind
strength and common envelope efficiency as hyperparameters $\bm{\lambda}$, and
random variables that affect certain stochastic process such as whether CO natal
kick unbinds a binary as $\bm{X}$. To avoid clutter in the following derivation,
we denote all the parameters related to mapping a particular progenitor system
into a GW event collectively as evolutionary parameters $\bm{\Theta}$, which
includes $\bm{\theta'}$, $\bm{\lambda}$, and $\bm{X}$.

@@

\subsection{\textbf{Binary evolution assumptions}}
\label{subsec:binaries}

@@

We apply the method described in section \textbf{\ref{subsec:stats}} to the first GW event GW150914 \citep{GW150914}.
We use the \texttt{Overall\_posterior} posterior samples for GW150914,
publicly available from the Gravitational wave Open Science Center \citep{LIGOScientific:2019lzm}. 
The parameters involved in the analysis and the range we allow for the inference are tabulated in table \ref{tab:parameters}.
The spin of BHs formed in isolated binaries depends strongly on the angular
momentum transport in massive stars which is still highly uncertain (see however \citet{Fuller2019, Bavera2020}),
therefore we only consider the two component masses in the observable space $\bm{\theta}$ in both the
root-finding stage and the MCMC stage.

For progenitor parameters $\bm{\theta'}$, we characterize each guess with
five parameters: the two component masses $M_{\rm 1,ZAMS}$ and $M_{\rm 2,ZAMS}$, the orbital period $t_{\rm orb}$,
the eccentricity $e$, and the metallicity $Z$ at ZAMS. Note that the progenitor formation redshift is not
obtained through the root-finding process because the formation redshift does not intrinsically affect
the evolution of the binary. Instead, we compute the time it takes for the binary to evolve from
formation to the merger and add this time to the lookback time when a GW event is observed and limit the total
time to merger to be less than $13.7\,\rm{Gyr}$.
Using the total lookback time, we can compute the redshift of the binary at ZAMS formation following the
Planck 2018 cosmological model \citep{Planck2018}.
The main benefit of post-processing the redshift in this way is that we do not rely on a particular assumption
of star formation rate (SFR) distribution when we solve for the posterior distribution in redshift.
This means we can put the prior on formation redshift (or in general, metallicity-redshfit distribution) in
by reweighting the posterior distribution according to a particular SFR model,
and therefore do not have to rerun the inference every time we change the SFR model. This is especially helpful
since SFR models are still highly uncertain but strongly impact the local merger rates of BBHs \citep[e.g. ][]{Broekgaarden2021}.

For hyperparameters $\bm{\lambda}$, we choose parameters that strongly affect
binary evolution in \texttt{COSMIC} for massive stars, including the common envelope
efficiency $\alpha$, the fraction of mass accreted onto the accretor during stable mass transfer $f_{\rm acc}$,
the critical mass ratio which determines whether Roche-overflow mass transfer
proceeds stably or produces a common envelope when the donor star is on the Hertzsprung gap
$q_{\rm crit,3}$, and the root-mean-square of the Maxwellian distribution
employed for CO natal kicks, $\sigma$. \textbf{We follow the critical mass ratio prescription 
described in \cite{Neijssel2019} for donors which fill their Roche lobes during
 other phases of stellar evolution.} 
For systems which begin a common envelope 
with a Hertzsprung-gap donor, we always assume the pessimistic outcome that the two 
stars merge since the donor has likely not formed a strong core-envelope boundary 
\citep{Ivanova2004, Belczynski2008}. \textbf{We fix other assumptions for how stellar and binary 
evolution proceed. In particular, we assume that stars lose mass through stellar winds 
in a metallicity-dependent 
fashion according to \cite{Vink2001,Vink2005}. We assume that binaries are rapidly circularized
during both stable Roche-overflow mass transfer and common envelopes.
We further assume that the CO formation mechanism follows the delayed 
prescription of \citet{Fryer2012} which allows neutrino-driven supernova explosions to 
occur from instabilities which grow on timescales longer than $250\,\rm{ms}$, 
and that natal kicks due to supernovae are reduced by the amount 
of mass which falls back onto the proto-CO
as described in \citet{Fryer2012}.}

We show a portion of the joint posterior of the progenitor parameters and
hyperparameters of GW150914 in figure \ref{fig:GW150914_posterior}. Each panel
is colored based on the comparison between progenitor parameters (blue),
hyperparameters (orange), or a mix of the two (green). We find that the choice
for $\sigma$ does not affect the formation of GW150914-like BBHs. This is not
unexpected since \textbf{the \citet{Fryer2012} delayed model can significantly reduce} the
natal kick strength at BH formation based on the amount of fallback onto the
proto-CO. In the case of GW150914-like BBHs, the BH progenitors are massive
enough that the fallback reduces the natal kick to zero. Because of this we do
not include $\sigma$ in Figure~\ref{fig:GW150914_posterior}. Similarly, we do
not include the ZAMS orbital period and eccentricity, which are correlated with
one another, but not strongly with the other progenitor parameters or
hyperparameters. We find strong correlations between the ZAMS masses, as well as
strong correlations between $M_{2,\rm{ZAMS}}$ and $f_{\rm{acc}}$, and
$f_{\rm{acc}}$ and $\alpha$. We show scatter plots of each of these combinations
in the upper-right inset of the figure.
